Complexity-stability relationships in disordered dynamical systems

### 2024-04-16 papers/response.md

## Referee B

### Major comments

- Point to Science for ecological justifications and in general to answer the first point.
- Point to Science paper for feasibility adding some discussion on this in the main text maybe.
- Maybe adjust something in the text to be more clear that the derivation of the results was obtained with rescaling and then we look at the non rescaled value of $\mu$ and $\sigma$.
- Answer to the reviewer that in this kind of conceptual work we are not necessarily interested in the realistic form of the SAD, indeed usual GLV papers have the Gaussian SAD prediction. In any case the model can still give non gaussian distribution in the case in which we use random growth rates etc... To conclude, the gaussian approximation is used only to more clearly present the results about complexity-stability but it is not necessary.

### Minor comments
- Just engage with the reviewer discussing mutualism etc, but not change anything necessarily.
- We stick to $N$, we don't want to relate to ecological literature, but more to complex systems. Moreover May used $N$ in his paper.
- Just answer something about all the changes that we made.

@@

---

# Response to reviewer

## Referee A

- *The claim is probably true, but unfortunately it does not describe a generic situation, because it is very hard to imagine that the exponents alpha, beta, gamma used in the model are identical for all populations. Perhaps the model would be more interesting if it included a spectrum of different exponents, described, for example, by a probabilistic law.*

...

- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*

We thank the referee for this comment, we modified Equations (7) accoridngly.

- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>. The self-consistency condition requires that the probability distribution (12) has exactly the same expected values. This should be discussed in detail.*

We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarifications.

In some cases, for example $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations. However in potential case of interests, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ we consider resulting in a distribution with a power-law tail with exponent $-3/2$, the moments diverge. We now modified the discussion removing the gaussian approximation and considering the introduction of a cut-off. The idea behind this cut-off can be described by the example below.

Consider, as an example which allows for a complete analytical treatment, the case of a power law distribution 
$$
P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
$$
defined from 1 to $\infty$ and with 
$$
\mathcal{Z}=\int_1^{\infty}x^{-\alpha}.
$$

Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behaviour of the sample mean 
$$
\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
$$
where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
$$
\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
$$
with the cut-off $\Lambda$ defined such that $\int_1^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\alpha=3/2$ we have 
$$
\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
$$
and therefore $\Lambda=N^2$. We have for the mean
$$
\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
$$
The result is plotted below, alongside sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

![](../plots/cut-off-justification.png)

## Referee B

We thank the referee...

- *Modelling Framework Justification:*

     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws. Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term). The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution. If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface. The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*

...

- *Feasibility Condition Neglected:*

    *The authors do not seem to consider the condition for the feasibility of the fixed points (x\* >= 0), which is critical. This oversight, especially evident in Equations 11 and 12, i.e., Eq. 12 or its Gaussian approximation are normalized from 0 or from -infty to infty? I think this is a very relevant issue calling into question the appropriateness of the results.*

...

- *Lack of Proper Scaling:*

    *There is no proper scaling of the interactions, so the thermodynamic limit is not well defined. And this is not discussed or compared with other models in the literature applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv preprint arXiv:2306.11031, 2023). Relatedly, the comparison with GLV with quenched disorder (e.g. “The properties of large random dynamical systems is often portrayed in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the interactions strengths with the system size.*

...

- *Gaussian Approximation:*

    *The justification for employing a Gaussian approximation is unclear, given that Equation 12 should be solved self-consistently. Ecologically, Gaussian species abundance distributions (SADs) are not meaningful due to the prevalence of heavy-tailed SADs in natural ecosystems.*

We thank the referee for this comment, we now removed the Gaussian approximation in the discussion and clarified how to treat the case in which self-consistently computing the moment in Equation (12) can be problematic, as explained in the answers to referee A.

- *Homogenous Case and Stability:*
     *The manuscript concludes that stability requires stronger self-interactions than cross-interactions when alpha equals beta. While this is true that self-interactions promote stability also for the Lotka-Volterra model, any negative mean species interaction (indicating mutualism) in the homogeneous case leads to instability, irrespective of system size. Thus, comparing this to the GLV model seems a stretch.*

...

- *Notation:*

     *Using S instead of N for species richness would align better with ecological conventions.*

...

- *Discussion Quality:*

    *Given the issues raised, the discussion section of the manuscript needs substantial improvement to convincingly argue the paper's contributions to the field. In light of these concerns, before the authors submit their manuscript elsewhere, I recommend to undertake significant revisions to address the weaknesses in their modelling framework, the rationale behind their approach, and the ecological implications of their findings*

...

### 2024-04-16 papers/submission/beyond-may.tex

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems.
\end{abstract}

@@

%Model
\paragraph*{}
\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a negative (resp.
positive) effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
We refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.

However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,Mazzarisi2023}.

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case, the first and second moments formally diverge.
This is not truly problematic because we are dealing with large, but \emph{finite} populations.

Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$, it diverges for $N\to\infty$. For finite $N$ we have $\bar{x}_N\neq\langle x_*\rangle=\infty$. What we really have is the convergence of $\bar{x}_N \approx \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ dependence on $N$ is such that $1-\int_0^{\Lambda}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$.

Following~\cite{Cui2020,Hatton2024} and assuming $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}\ll \tilde{\mu}  \langle x_* \rangle$, we can expand the solution~\eqref{eq: cavity solution} to first order in $\tilde{\sigma} \sqrt{\langle x_*^2\rangle}$, and approximate~\eqref{eq: dist general} with a Gaussian distribution with moments $\langle x_*\rangle=\tilde{\mu}^{-2/3}$ and $\langle x_*^2\rangle=\tilde{\mu}^{-4/3}/(1-4\tilde{\mu}^{-2}\tilde{\sigma}^2)$.

The result is plotted against simulations in Fig.~\ref{fig: cavity sol.}. We show both the exact solution~\eqref{eq: dist general} (where integration for the moments has been performed up to a cut-off above which no equilibrium value should be found) and the Gaussian approximation. Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

@@

In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading). Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region. This behavior corresponds to the ``anti-May" phase defined previously.

@@

A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interactions) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
All the models in this class are destabilized by diversity.
The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.

>>>>

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems:
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems.
\end{abstract}

@@

%Model
\paragraph*{}
\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) - \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-dynamics of a population $i$ (growth and self-regulation), while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}>0$, such that it implies a negative effect of $j$ on the growth of $i$. Non-zero diagonal elements $A_{ii}$ accounts for additional self-interactions with the same functional form as cross-interactions. 
We refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models of complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
The classic generalized Lotka-Volterra (GLV) competition model studied by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear.

However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case, the first and second moments formally diverge. It is not uncommon, for example in ecology, to find empirical species abundance distributions characterized by heavy-tails, potentially with diverging moments.
This is problem can be overcome considering that we are dealing with large, but \emph{finite} populations. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$.

Following, e.g.,~\cite{Cui2020,Hatton2024} we can compute self-consistently $\Lambda$, $\langle x_*\rangle_{\Lambda}$ and $\langle x_*^2\rangle_{\Lambda}\equiv\int_0^{\Lambda}dx_*x_*^2P(x_*)$. The resulting distribution is plotted against simulations in Fig.~\ref{fig: cavity sol.}. Simulations correspond to $A_s=\mu$, showing that neglecting the $A_s$ term in analytical expressions does not introduce a significant error.

@@

In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x_*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading). Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region. This behavior corresponds to the ``anti-May" phase defined previously.

@@

A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interaction strength) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
All the models in this class are destabilized by diversity.
The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.
